---
title: "1 - Introduction "
author: "Cory Costello"
date: "September 25, 2017"
output: 
  md_document:
    preserve_yaml: true
order: 1
---
#What the F#$% is R and how do I get started with it?

##Installing R and R Studio 

First, you will need to download R if you haven't already. You can do that at this link: <https://www.r-project.org/>

Second, you'll want to download R studio. You can use R without R studio, but you probably don't want to. R studio is an integrated development environment (IDE; <https://en.wikipedia.org/wiki/Integrated_development_environment>). R studio can be downloaded at the following link: <https://www.rstudio.com/>

Okay, that is all you need to start your gloRious adventuRe.

##How do I R?!
### Using RStudio
There are four main panels in the RStudio window.  

The **top right** panel is the **workspace**. Here you will find a list of all objects currently in memory.  

The bottom right window displays files in your working directory, plot output, available packages, and help documentation.  

The **bottom left** panel is the **console**. You can type code directly into the console and hit enter to execute it.  

The **top left** panel is the **code editor**. This is where you will write most of your code. In the code editor, you can write multiple lines of code and run chunks or the full script. The toolbar at the top of the panel allows you to interact with the code in the editor in some important ways. Hover your mouse pointer over the icons to see what each button does.  

There are also shortcuts that allow you to use the toolbar functions without ever taking your fingers from the keyboard. Here are some of the most commonly used:  

Function | Keys
---------- | --------
Run highlighted code | `Cmd+Enter`
Insert <- | `option+-`
(Un)comment lines | `Cmd+Shift+C`
Knit document | `Cmd+Shift+K`
Run from start to current line | `Cmd+Option+B`
Keyboard shortcut quick reference | `Option+Shift+K`

For a more complete list, check out the RStudio IDE Cheat Sheet. Find it online by going to `Help>Cheatsheets>RStudio IDE Cheatsheet`, or by clicking the following URL <https://www.rstudio.com/wp-content/uploads/2016/01/rstudio-IDE-cheatsheet.pdf>  

Another important shortcut is `Tab` to autocomplete functions or file names.  

For a more complete list, check out the [RStudio IDE Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2016/01/rstudio-IDE-cheatsheet.pdf)


## Getting help 
Check out the [r-project site](https://www.r-project.org/help.html) for more on help documentation.  

#What is R?
R is a programming language designed for statistical computing. So, it is (in many ways) more similar to another programming language (e.g., Python) than it is to statistical software you may have experience with (e.g., SPSS). If you have no experience with programming languages and have only ever used statistical software (e.g., SPSS) for your stats then you will probably feel overwhelmed and confused by R for a little while. That's fine and normal. Speaking for myself, I think I installed and opened R studio about a half dozen times over the course of a year before I ever used it becuase I had no idea where to get started. Hopefully this (and the next few weeks of the Data Science Club) helps you with that.

How do we use R. Lots of ways! For starters, it's like a calculator. Sort of like the graphing calculators that you may have been fascinated by at 16 (or at least I was for some reason). Because it's like a calculator, we can use it like one.
```{r R is a fu$%ing calculator}
5+6
```
Okay, but now we all have calculators in our phones, so this isn't that cool. What else can R do?

Lots of stuff. Almost all of it revolves around *assigning* and then later *calling* and *manipulating* objects. What is an object? It's just R's way of representing information. And then we can use that information later. 


##Objects
How do we get R to represent some information as an object? We tell it to, using the assignment operator '<-'. Whatever we put on the left side of the arrow becomes the name of the object and whatever is on the right side of the arrow becomes the definition of that object. For example, we may want to store what we had done earlier into an object.

```{r objects}
spinal_tap_amps <- 5 + 6
```
Now, we can tell R to do something with this object. For example, we could get its sqaure root:

```{r using objects 1 - math}
sqrt(spinal_tap_amps)
```
Or we could use it in a logical statement

```{r using objects 2 - logical}
if (spinal_tap_amps == 11) {
  print("But these go up to 11")
}

# And we could do this with other objects:
other_amps <- 10

if (other_amps == 11) {
  print("But these go up to 11")
}
```
I know the above had some new stuff in it (like wtf is with the curly brackets), but it's just to demonstrate that we can do more with objects than just math.

We now have 2 different objects: spinal_tap_amps and other_amps. We can put these together into an object called amps. We'll do this like so:

```{r combining objects}
amp_volumes <- c(spinal_tap_amps, other_amps)
amp_volumes
```
What happens if we take the sqrt of our new object?

```{r operating on our new object}
sqrt(amp_volumes)
```
Would you look at that! It gave us the square roots of each of the values in our new object. How did it know?

It knows because when we took two single-element objects and put them together, R recognized this new object as a *vector*. When we tell R to perform an operation on a vector, it will tend to perform that operation on each element. This will not always be the case, as certain functions will tell R to do something else, but it is generally the case.

Now I know what you're thinking. I don't work with vectors I work with datasets! Just like we combined individual elements into a vector, we can combine vectors into a dataset. Since we've been working with the volume capacity of guitar amps (in case this hasn't been obvious, this has all been a reference to spinal tap; see <https://en.wikipedia.org/wiki/Spinal_Tap_(band)>), let's make a vector for the names of the bands we're considering.

```{r character vector}
Bands <- c("Spinal Tap", "Everyone Else")
```

Now let's put that and our vector of numbers together

```{r combining vectors 2}
amp_volume_dataset <- cbind(amp_volumes, Bands)
``` 
And voila! We have a very simple dataset. Notice in the environment, R studio knows that this new object is data rather than values (values includes single objects/elements and vectors).

R has a few different formats for data. We'll likely talk more about these in later weeks as they become relevant. But check this out anyway:
```{r changing data format}
amp_volume_dataset <- as.data.frame(amp_volume_dataset)
```
See, now it's a data frame. We could also change it to a data table (with as.table()) or other formats.

The above illustrates another thing about r. If you use the same name for a new object, it will overwrite the old object. Now we can't get our old unformatted data object. We will have to make it again if we want to:
```{r returing to unformatted}
amp_volume_dataset <- cbind(amp_volumes, Bands)
```
And if we wanted both, we can just assign the dataframe to a new object

```{r dataframe too}
amp_volume_dataframe <- as.data.frame(amp_volume_dataset)
```

##Functions

Now that you know about *objects*, you'll want to know about *functions*. In R, we use functions to do stuff to objects. Functions take in some input, perform some operation, and then provide some output. We've already covered several functions:
*c() is a function for putting things together 
+it stands for concatenate(mneumonic: it also shares an initial with combine)
+it takes objects as the input, combines them, and provides the combined object as an output
*sqrt() is a function
+it takes numeric objects as inputs (elements, vectors, matrices), calculates square-root of each, and outputs the calculated square root.
*as.data.frame() is a function
+it takes some object and formats it as a dataframe
*print() is a function
+it takes some object as input, and the output is that object printed to the console
*cbind() is a fucntion
+it takes at least two objects, and combines them column-wise

Some of the functions that you'll use are already in your install of R. You can write your own functions (more on this later on) or you can use one's others have written, like sqrt(). 

Say we are confused about how a function works. We can google it (and you should) and we can also use R's built in help. To access the help file of a function, you enter a '?' followed by the package name:
```{r help function easier}
?sqrt
```

The help documentation can be pretty confusing if you aren't quite familiar with the package/function. The examples (which are typically at the bottom) can be very helpful, even for beginners. And some are simpler than others:

This one basically just says it computes square root of x, where x is a vector or array. It also tells you (subtly) what the function works with; square root expects a numerical vector (including a vector of 1, or an element). If we give it a vector of non-number objects (like the Bands list above), it won't work. Sometimes checking the help documentation can tell you if you've given the function the wrong type of object.

## Packages

Often times, functions are part of *packages* or *libraries* in R. These are usually made by other people and have lots of functions you might find useful. One that we will start using soon is called the tidy verse. It's actually a set of packages that the authors put together (you can also get each of the packages separately, butyou don't need to). To use a new package, you'll first need to install it. Let's install the tidyverse:

```{r installing a package, eval=FALSE, echo=TRUE}
#install.packages("tidyverse")
```
We just used the install.packages() function. This goes to cran, and downloads the package into our user library (which is where we store all of the packages we install). It requires you to enter the package name in quotes, like we did above.

How does `install.packages()` know where to download the package from, and where does it go? Check its documentation with `?install.packages`. In the list of arguments we see `lib` and `repos = getOption("repos")`. Further down we see that `lib` defaults to the first element of `.libPaths()`. What do we get when we run those commands?
```{r package details, eval=TRUE, echo=TRUE}
getOption("repos")
.libPaths()
```

Most packages we use will come from cran, and if you ever need to modify or find package files for any reason, there they shall be.

Now we have the package tidyverse, and can use it whenever we want. I'm so excited that I'm going touse one of the packages within the tidyverse called ggplot to make a simple plot of our spinal tap influenced dataset!!!!!

```{r GGPlot, eval=FALSE, echo=TRUE}
ggplot(data=amp_volume_dataframe, aes(x=Bands, y = amp_volumes))+
  geom_col()
```
OMG, it didn't work!!! Why?!?! I guess I'll have to go back to my unsatisfying life in SPSS.

Wait, its no big deal, we just forgot one simple step. In order to use a package, we have to call the library, or tell R that we're going to be using this library (we're basically telling it to keep it in its working memory). We load libraries into our environment with the library() function. We don't need to put quotes (ie. make it a string) around the name in the library function (side note: you will make the mistake of quoting things in library and not quoting things in install.packages upwards of 100 times). This puts the library in the list of packages to search. For example, compare the output of `search()` before and after loading a library:

```{r search paths}
search()
library(tidyverse)
search()
```

Importantly, you have to call all of the libraries you will use each time you start a new session in R. You may want to get in the habit of putting a bunch of library calls at the top of your script.

```{r call a library}
library(tidyverse)
```
And now let's try to use ggplot
```{r GGPlot works!}
ggplot(data=amp_volume_dataframe, aes(x=Bands, y = amp_volumes))+
  geom_col()
```
It worked! We'll spend some time talking about ggplot (led by Jonny) a little later on, so don't worry about what that code means. Soon you'll know.

What happens when two packages have functions with the same name? for example, both dplyr and plyr have functions named "summarise" that behave slightly differently. We can solve these "namespace conflicts" by being specific about where we load the library in our search paths. If we want to load the functions from plyr, but keep all our existing references to dplyr, we could do something like this (although if you've already loaded tidyverse, this won't do anything because these packages are loaded by tidyverse)
```{r ordered library load}
library(plyr, pos=100) # We can put it at an arbitrarily high number, it will be after everything but the base functions
library(dplyr)
```

Or, we could make our code more explicit. We can directly reference a package's functions without loading the library by using `::`, for example with 
```{r direct function reference}
ggplot2::ggplot(data=amp_volume_dataframe, ggplot2::aes(x=Bands, y = amp_volumes))+
  ggplot2::geom_col()
```

That is all for this basic overview of R. Hopefully it has started to make youR adventuRe moRe toleRable!