---
title: "2.1 - Objects"
author: "Jonny Saunders"
date: "October 5, 2017"
output: 
  md_document:
    preserve_yaml: true
    toc: true
    toc_depth: 2
order: 2
---

# What are Objects?

When using R, only objects in active memory are manipulated, not the data on the disk. So your data files are always available in their raw format, which means your analyses can be reproduces and altered without ever touching the raw data files.   

Objects are, roughly, data (or more generally a stored state) that know what it can do. 

We know what happens when we put this troublesome one between numbers

```{r}
1 + 1 # ud better b sitting down
```

But it's less clear what it means to `+` letters

```{r, error=TRUE}
"a" + "b"
```

Let's see what `typeof` variables `1` and `"a"` are  :
```{r}
typeof(1)
typeof("a")
```
>(note this is a little misleading, `typeof` determines the base object class that an R object is stored as. All R objects are composed of base objects, we'll get to the types of objects in the next section)


## Object terminology
A **class** is the description, or 'blueprint' of how individual **objects** are made, including their **attributes** - which data should be kept and what it should be named, and **methods**, the functions that they are capable of calling on their stored data or attributes. 

See this example class ():

> do the example class


*classes define the behavior of objects
** Describe attributes and relationship to other classes
** Methods are functions that depend on the class of their input. try methods(object) (`?methods`)



# Objects in R
>"In R functions are objects and can be manipulated in much the same way as any other object." - *R language guide 2.1.5*

>"S3 objects are functions that call the functions of their objects" - *Also R*


R has three object-oriented systems and the base types.

* **S3 - "Casual objects":** Objects that use **generic functions**. Rather than asking an object what method to call, the generic function decides which method to call based on the argument classes. Functions contain the **UseMethod("function_name", object)** function (see `?UseMethod`), which searches the object's namespace for "function_name" and passes computation to it.
** Find example of S3 object

* **S4 - "Formal objects":** Formal classes with inheritance and means by which methods can be shared between classes.

* **Reference classes:** Objects that use **message passing** - or the method 'belongs to' the class rather than a function. This is the common `dataframe$column_name` syntax.

* **Base types:** Low-level C types. Build the other object systems.

R has a useful package `pryr` for inspecting objects and other meta-linguistic needs. Let's get that now.
```{r}
# install.packages("pryr")
```

We can query any object's type with pryr's `otype`
```{r}
pryr::otype(c(1,2,3,4,5)) # A vector is a base object
pryr::otype(data.frame(x=c(1,2,3,4,5))) # A dataframe is an S3 object

```




# S3 Objects
S3 objects "belong to" functions, S4 objects "have" functions (methods). 

S3 objects are defined by a series of functions that themselves contain the `UseMethod()` function - this is described briefly above, try `?UseMethod` for more detail. These functions extend the generic function, typically using the syntax `generic.class()` as in the case of `mean.Date()` for taking the mean of dates. One can list the methods of an S3 class with `methods()`

By default, the source code of S3 methods is not visible to R, one can retreive it with `utils::getS3method``

The `plot` base function is an s3 generic method.
```{r}
pryr::ftype(plot) # get a function's type
```

By default, if the first argument is a base type compatible with being points on a scatterplot, the actual function that is called is `plot.default`, whose source behaves like you'd expect:

```{r}
plot.default
```

If the first argument to `plot` has its own `plot` method (ie. that it is exported by the object's package namespace, more about this in section 5), that function is called instead. That's why 

```{r}
aq <- datasets::airquality
plot(lm(Ozone ~ Month, data=aq))
```

is different than this nonsensical model

```{r}
plot(lme4::lmer(Ozone ~ 0 + (Day | Month), data=aq))
```

## Extending S3 Objects
> http://adv-r.had.co.nz/OO-essentials.html "Creating new methods and generics"

```{r, error=TRUE}
f <- function(x){
    UseMethod("f")
}
f.a <- function(x) "Class a"
a <- list()
class(a)
class(a) = "a"
class(a)

f(a)

b <- list()
f(b)

```



# S4 Objects


# Base Types
http://adv-r.had.co.nz/Data-structures.html#data-structures

# References
* http://manuals.bioinformatics.ucr.edu/home/programming-in-r#TOC-Object-Oriented-Programming-OOP-
* http://www.stat.ucla.edu/%7Ecocteau/stat202a/resources/docs/S4Objects.pdf
* http://adv-r.had.co.nz/OO-essentials.html


-------------------

```{r}
# S4 Methods are stored in environments 
nM <- asNamespace("Matrix")
sort(grep("^[.]__T__", names(nM), value=TRUE))
meth.Ops <- nM$`.__T__Ops:base`
head(sort(names(meth.Ops)))
```

