---
title: "2.1 - Objects"
author: "Jonny Saunders"
date: "October 5, 2017"
output: 
  md_document:
    preserve_yaml: true
    toc: true
    toc_depth: 2
order: 2
---

# What are Objects?

Objects are, roughly, data (or more generally a stored state) that knows what it can do. 

We know what happens when we put this troublesome `+` guy between numbers

```{r}
1 + 1 # ud better b sitting down
```

But it's less clear what it means to `+` letters

```{r, error=TRUE}
"a" + "b"
```

Let's see what `typeof` variables `1` and `"a"` are  :
```{r}
typeof(1)
typeof("a")
```
>(note this is a little misleading, `typeof` determines the base object class that an R object is stored as. All R objects are composed of base objects, we'll get to the types of objects in the next section)


## Object terminology
A **class** is the description, or 'blueprint' of how individual **objects** or **instances** are made, including their **attributes** - which data should be kept and what it should be named, and **methods**, the functions that they are capable of calling on their stored data or attributes. Objects can have a nested structure, and sub-classes can **inherit** the attributes and methods of their parent classes.

For example: As a class, trucks have attributes like engine_size, number_of_wheels, or number_of_jumps_gone_off. Trucks have the method go_faster(), but only individual instances of trucks can go_faster() - the concept/class of trucks can't. As a subclass, monster_trucks also have the attributes engine_size, etc. and the method go_faster(), but they also have additional attributes like mythical_backstory and methods like monster_jam().


*classes define the behavior of objects
** Describe attributes and relationship to other classes
** Methods are functions that depend on the class of their input. 



# Objects in R
>"In R functions are objects and can be manipulated in much the same way as any other object." - *R language guide 2.1.5*

>"S3 objects are functions that call the functions of their objects" - *Also R*


R has base types and three object-oriented systems.


* **Base types:** Low-level C types. Build the other object systems.

* **S3 - "Casual objects":** Objects that use **generic functions**. Rather than asking an object what method to call, the generic function decides which method to call based on the argument classes. Functions contain the **UseMethod("function_name", object)** function (see `?UseMethod`), which searches the object's namespace for "function_name" and passes computation to it.
** Find example of S3 object

* **S4 - "Formal objects":** Formal classes with inheritance and means by which methods can be shared between classes.

* **Reference classes:** Objects that use **message passing** - or the method 'belongs to' the class rather than a function. This is the common `dataframe$column_name` syntax.

R has a useful package `pryr` for inspecting objects and other meta-linguistic needs. Let's get that now.
```{r}
# install.packages("pryr")
```

We can query any object's type with pryr's `otype`
```{r}
pryr::otype(c(1,2,3,4,5)) # A vector is a base object
pryr::otype(data.frame(x=c(1,2,3,4,5))) # A dataframe is an S3 object

```

# Base Types
Every R object is built out of basic C structures that define how it is stored and managed in memory. 

This table from [Advanced R - Data Structures](http://adv-r.had.co.nz/Data-structures.html#data-structures) summarizes them:

 | Homogenous data | Heterogenous data
---------------- | ------------------
1 Dimensional | Atomic Vector | List
2 Dimensional | Matrix | Data frame
nd | Array

## Atomic Vectors

## Lists & Matrices

## Data Frames


# S3 Objects
S3 objects "belong to" functions, S4 objects "have" functions (methods). S3 classes don't really "exist," but are assigned as an object's "class" attribute.

```{r, echo=TRUE}
x <- 1
attr(x, "class")
class(x) <- "letters"
attr(x, "class")
```


S3 objects are defined by a series of functions that themselves contain the `UseMethod()` function - this is described briefly above, try `?UseMethod` for more detail. These functions extend the generic function, typically using the syntax `generic.class()` as in the case of `mean.Date()` for taking the mean of dates. One can list the objects that have a generic method, and the methods that an object has with `methods()`

```{r}
methods(mean)

methods(class="Date")
```


By default, the source code of S3 methods is not visible to R, one can retreive it with `utils::getS3method``

The `plot` base function is an s3 generic method.
```{r}
pryr::ftype(plot) # get a function's type
```

By default, if the first argument is a base type compatible with being points on a scatterplot, the actual function that is called is `plot.default`, whose source behaves like you'd expect:

```{r}
plot.default
```

If the first argument to `plot` has its own `plot` method (ie. that it is exported by the object's package namespace, more about this in section 5), that function is called instead. That's why 

```{r}
aq <- datasets::airquality
plot(lm(Ozone ~ Month, data=aq))
```

is different than this nonsensical model

```{r}
plot(lme4::lmer(Ozone ~ 0 + (Day | Month), data=aq))
```

## Example: Extending S3 Objects
> http://adv-r.had.co.nz/OO-essentials.html "Creating new methods and generics"

```{r, error=TRUE}
pryr::ftype(mean) # mean is an s3 generic function


x <- 1
class(x) <- "just_one"

# We give our "just_one" class a mean method:
mean.just_one <- function(x, ...) print("that's just a one you maniac")

# Mean behaves like it should for numbers and lists of numbers
mean(1)
mean(c(1,1.5))

# Other objects have their own mean() method
methods(mean)

# like Date objects
dates <- c("01jan2000","15jan2000")
attr(dates,"class")
mean(dates) # Don't work


# turn it into "Date" object
dates <- as.Date(dates, "%d%b%Y") # base has a set of "as" methods to convert types
attr(dates,"class")
mean(dates) # will call its method

mean.Date(dates) # which can also be called directly
```



# S4 Objects

# References
* http://manuals.bioinformatics.ucr.edu/home/programming-in-r#TOC-Object-Oriented-Programming-OOP-
* http://www.stat.ucla.edu/%7Ecocteau/stat202a/resources/docs/S4Objects.pdf
* http://adv-r.had.co.nz/OO-essentials.html


-------------------

```{r}
# S4 Methods are stored in environments 
nM <- asNamespace("Matrix")
sort(grep("^[.]__T__", names(nM), value=TRUE))
meth.Ops <- nM$`.__T__Ops:base`
head(sort(names(meth.Ops)))
```

